---
title: "README"
author: "Piyal Karunarathne"
date: "6/17/2021"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start -->
[![Github All Releases](https://img.shields.io/github/downloads/milesilab/BioRssay/total.svg)]()
[![R-CMD-check](https://github.com/milesilab/BioRssay/workflows/R-CMD-check/badge.svg)](https://github.com/milesilab/BioRssay/actions)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![status](https://joss.theoj.org/papers/73183faaf2c750eaaf0adaecc75d1f8f/status.svg)](https://joss.theoj.org/papers/73183faaf2c750eaaf0adaecc75d1f8f)
<!-- badges: end -->

**BioRssay** will help you to analyze mortality data from bioassays of any kind (e.g.,insecti/pesti/herbi-cides, antibiotics, etc.).

```{r,echo=FALSE}
library(BioRssay)
data(bioassay)
transd<-probit.trans(bioassay$assay2)
data<-transd$tr.data
strains<-levels(data$strain)
mort.plot(data,strains)
```

***For a comprehensive tutorial on the package, go to <https://milesilab.github.io/BioRssay/> and navigate to "Get started" where all the functions and usage are explained with ample examples.***

**Installation**
```{r,results = "hide",eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) 
    install.packages("devtools") 
devtools::install_github("milesilab/BioRssay")

```

Please don't forget to cite us if you use the package.

**How to cite**

*Publication to be added

